### 2024-06-25 basic-statistics.qmd
20240624 - p-value misconceptions
20240624 - tweak figure

For a test of whether two variables are associated, the *p*-value evaluates the likelihood that you would observe an association as strong or stronger than the one you observed between the groups if there were no association between the variables.
For instance, when evaluating whether number of carries is <u>positively</u> associated with injury likelihood, and you observed a correlation coefficient of $r = .25$ between number of carries and injury likelihood, the *p*-value evaluates the likelihood that you would observe a correlation as strong or stronger than $r = .25$ between the variables if number of carries is not associated with injury likelihood.

@@

>>>>

For a test of whether two variables are associated, the *p*-value evaluates the likelihood that you would observe an association as strong or stronger than the one you observed between the groups if there were no association between the variables, as depicted in @fig-pValues.
For instance, when evaluating whether number of carries is <u>positively</u> associated with injury likelihood, and you observed a correlation coefficient of $r = .25$ between number of carries and injury likelihood, the *p*-value evaluates the likelihood that you would observe a correlation as strong or stronger than $r = .25$ between the variables if number of carries is not associated with injury likelihood.

```{r}
#| layout-ncol: 2
#| fig-cap: "*p*-Values"
#| fig-subcap: 
#|   - "What is the probability my data would look like this..."
#|   - "...if in the population the association was really this?"
#| label: fig-pValues
#| code-fold: true

set.seed(52242)
correlations <- data.frame(criterion = rnorm(2000))
correlations$sample <- NA
correlations$sample[1:100] <- complement(correlations$criterion[1:100], .9)
correlations$population <- complement(correlations$criterion, 0)

ggplot2::ggplot(
  data = correlations,
  mapping = aes(
    x = sample,
    y = criterion
  )
) +
  geom_point() +
  scale_x_continuous(
    limits = c(-3.5,3)
  ) +
  ggplot2::labs(
    x = "Predictor Variable",
    y = "Outcome Variable",
    title = "What is the probability my data would look like this..."
  ) +
  ggplot2::theme_classic(
    base_size = 16) +
  ggplot2::theme(
    legend.title = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank())

ggplot2::ggplot(
  data = correlations,
  mapping = aes(
    x = population,
    y = criterion
  )
) +
  geom_point() +
  scale_x_continuous(
    limits = c(-2.5,2.5)
  ) +
  ggplot2::labs(
    x = "Predictor Variable",
    y = "Outcome Variable",
    title = "...if in the population, the association was really this:"
  ) +
  ggplot2::theme_classic(
    base_size = 16) +
  ggplot2::theme(
    legend.title = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank())
```

@@

That is, the *p*-value is <u>not</u>:

- the probability that the effect was due to chance
- the probability that the null hypothesis is true
- the size of the effect
- the importance of the effect
- whether the effect is true, real, or causal
